---
output:
  md_document:
    variant: markdown_github
---

# Reproducible Research Week 2: Assignment 1

## Step 1: Code for reading in the dataset and/or processing the data
Initiate required Libraries 
```{r}
library(lattice)
```

Creating a master file for reference
```{r}
dfMaster <- read.table("activity.csv",
                       header=TRUE,
                       sep=",",
                       stringsAsFactors = FALSE,
                       colClasses = c("numeric","Date","numeric")
                      )
```

Creating the working dataframe
```{r}
df <- dfMaster
```

## Step 2: Histogram of the total number of steps taken each day
```{r}
TotalStepsByDay <- aggregate(df$steps,list(date=df$date),sum,na.rm=TRUE)
xAxisBreakdown = seq(from=0,to=25000,by=2000) 
hist(TotalStepsByDay$x,
     breaks = xAxisBreakdown,
     main="Frequency of Total Steps per Day",
     col="blue",
     xlab="Steps",
     ylab="Days",
     xaxt="n")
axis(side=1,at=xAxisBreakdown,labels=xAxisBreakdown)
```

## Step 3: Mean and median number of steps taken each day

Calculate and print Mean number of steps per day
```{r}
stepMean <- mean(TotalStepsByDay$x,na.rm=T)
stepMedian <- median(TotalStepsByDay$x,na.rm=T)
print(paste("The Mean number of steps per day is",round(stepMean,1)))
```

Calculte and print Median number of steps per day
```{r}
print(paste("The Median number of steps per day is",round(stepMedian,1)))
```

## Step 4: Time series plot of the average number of steps taken

Convert the 5-minute 24-hour clock intervals into a time class
The hours of the interval are equal to the first two digits (interval divided by 100, with remainder tossed)
```{r}
intHours <- df$interval %/% 100
```
Now add a leading 0 when the hours are one digit
```{r}
intHours <- ifelse(intHours < 10,paste("0",intHours,sep=""),intHours)
```
The minutes of the interval are equal to the second two digits (the remainder after dividing by 100)
```{r}
intMinutes <- df$interval %% 100
```
Now add a leading 0 when the minutes are one digit
```{r}
intMinutes <- ifelse(intMinutes < 10,paste("0",intMinutes,sep=""),intMinutes)
```
Now put the minutes and hours together and convert to a time with strptime()
```{r}
intTime <- paste(intHours,":",intMinutes,sep="")
intTime <- strptime(intTime,format="%H:%M")
```
Now add the time variable back into the original data set
```{r}
df <- cbind(df,intTime)
```
We now have the time in the data frame as a time, so we can compute the mean number of steps for each time interval and plot the result
```{r}
SPI <- aggregate(df$steps,list(intTime=df$intTime),mean,na.rm=TRUE)
plot(SPI$intTime,SPI$x,
     type = "l",
     main = "Average Steps per Interval",
     xlab = "Interval",
     ylab = "Average Steps")
```

## Step 5: The 5-minute interval that, on average, contains the maximum number of steps
```{r}
MaxStepAvg <- max(SPI$x)
IntervalWithMaxStepAvg <- SPI$intTime[SPI$x == MaxStepAvg]
```
The highest average is equal to `r MaxStepAvg`.

## Step 6: Code to describe and show a strategy for imputing missing data

### Part A: Calculate and report the number of missing values in the dataset.
```{r}
countNAs <- sum(is.na(df$steps))
```
The total number of missing values are `r countNAs`

### Part B: Create a dataset that replaces NA values with the mean for that interval.
Rename column "x" in data frame SPI to what it actually represents
```{r}
names(SPI)[names(SPI)=="x"] <- "avgIntervalSteps"
```
Now merge the average back into the dataframe by intTime
```{r}
dfWithAvg <- merge(x=df,y=SPI,by="intTime",all.x=TRUE)
```
Reorder the new dataframe to preserve the date, interval ordering of the original file
```{r}
dfWithAvg <- dfWithAvg[order(dfWithAvg$date,dfWithAvg$intTime),]
```
Create another column that uses the steps, if available, and the avgIntervalSteps otherwise.
```{r}
dfWithAvg$imputedSteps <- ifelse(is.na(dfWithAvg$steps), 
                                 dfWithAvg$avgIntervalSteps,
                                 dfWithAvg$steps)
```

## Step 7: Histogram of the total number of steps taken each day after missing values are imputed
```{r}
TotalStepsByDayImputed <- aggregate(dfWithAvg$imputedSteps,list(date=dfWithAvg$date),sum,na.rm=TRUE)
xAxisBreakdown = seq(from=0,to=25000,by=2000) 
hist(TotalStepsByDayImputed$x,
     breaks = xAxisBreakdown,
     main="Frequency of Total Steps (imputed) per Day",
     col="blue",
     xlab="Steps",
     ylab="Days",
     xaxt="n")
axis(side=1,at=xAxisBreakdown,labels=xAxisBreakdown)
```
Recalculate mean
```{r}
stepMeanImputed <- mean(TotalStepsByDayImputed$x,na.rm=T)
stepMedianImputed <- median(TotalStepsByDayImputed$x,na.rm=T)
print(paste("The Mean number of imputed steps per day is",round(stepMeanImputed,1)))
```
Recalculate median
```{r}
print(paste("The Median number of imputed steps per day is",round(stepMedianImputed,1)))
```

## Step 8: Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
```{r}
dfWithAvg$weekday <- weekdays(dfWithAvg$date)
dfWithAvg$weekendFlag <- ifelse(dfWithAvg$weekday=="Saturday" | 
                                  dfWithAvg$weekday=="Sunday","Weekend","Weekday")
```
Compute the average steps / day for weekend and weekdays
```{r}
SPI2 <- aggregate(dfWithAvg$imputedSteps,list(
                  intTime=dfWithAvg$intTime,weekendFlag=dfWithAvg$weekendFlag),
                  mean,na.rm=TRUE)
```
Before plotting, set up the sequence to appear on the x-axis
```{r}
xn <- seq(min(dfWithAvg$intTime),max(dfWithAvg$intTime),by="4 hour")
```
Draw the plot with the x-axis in HH:MM format
```{r}
xyplot(x~intTime|weekendFlag,
       data=SPI2,
       type="l",
       layout=c(1,2),
       xlab = "Time Interval (24-hour clock)",
       ylab = "Average Steps",
       main = "Average (imputed) Steps per Day - Weekend vs Weekday",
       scales=list(
         x=list(
           at=xn,
           labels=format(xn,"%H:%M")
         )
       )
)
```
